Probability Model

### 2019-07-17 NAR_DRAFT.tex
make a start fixing the methods/probability model section

Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.

\section{MATERIALS AND METHODS}

\subsection{Base probability model}

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
\begin{equation*}
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
\end{equation*}

where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as

\begin{equation*}
  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
    \begin{array}{cr}
      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
      \frac{10^{-q_i/10}}{3} & otherwise.
    \end{array}
    \right.
\end{equation*}

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or \textrm{t\_lod\_fstar} depending on the version.
By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.

\subsection{Estimation of the mutation signature.}

While variant calling algorithms typically assume a constant probability of mutation at every site in the genome, work by Alexandrov and others show that the random mutation generating process actually varies from site to site in a nucleotide context specific manner.
We develop a model of the prior probability of mutation to allele $m$ conditional on the observed genomic context $\mathrm{p}(m,M \mid C)$, and demonstrate an empirical Bayes method for computing this probability from MuTect output.
The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
$$
\mathrm{p}(m,M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(M \mid C) = \mathrm{p}(m \mid C) \mathrm{p}(C \mid M)\frac{p(M)}{p(C)}
$$
since the probability of a mutation at a site and the probability that it is to allele $m$ are independent conditional on the context.
Here $\mathrm{p}(M) = \mu$ as above, and the empirical distribution of contexts $\mathrm{p}(C)$ is the fraction of the genome made up of each context.
We model $\mathrm{p}(C \mid M)$ as a multinomial distribution with parameter $\boldsymbol{\pi} = \{\pi_i\}, i = 1\dots96$.
Mutations are drawn from this multinomial distribution such that $\mathrm{p}(C = i \mid M) = \pi_i$.
The final quantity $\mathrm{p}(m \mid C)$, the probability of mutation to $m$ given a particular three letter context, is a function of $\boldsymbol{\pi}$. 
We are left to estimate only the vector of probabilities $\boldsymbol{\pi}$.

$$
\begin{aligned}
C \mid M,\boldsymbol{\pi} & \sim \textrm{Multinomial}(\boldsymbol{\pi})  \\
\boldsymbol{\pi} \mid \boldsymbol{\alpha} &\sim \textrm{Dirichlet}(\boldsymbol{\alpha}).
\end{aligned}
$$

The posterior distribution of $\boldsymbol{\pi}$ is $\boldsymbol{\pi} \mid C,\boldsymbol{\alpha} \sim \textrm{Dirichlet}(\mathbf{C} + \boldsymbol{\alpha})$, where $\mathbf{C} = (C_1,\dots,C_{96})$ are the counts of mutations present in the tumor for each of the 96 contexts.
We compute an empirical bayes estimate of $\boldsymbol{\pi}$ by choosing $\mathbf{C}$ as the set of mutations assigned a TLOD by MuTect above some threshold, which we choose as 10.
We show through extensive simulation that our estimate of $\boldsymbol{\pi}$ converges quickly its true simulated value.

Returning to the model above, we can calculate the log posterior odds in favor of $\mathbf{H_1}$ as

$$
  \textrm{log}_{10} \left(\frac{(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})\mathrm{p}(m,M \mid C)}{(\mathcal{L}_{\nu=0,m}(\mathbf{x})(1-\mathrm{p}(m,M \mid C))} \right) = \textrm{TLOD} + \textrm{log prior odds},
$$

and the posterior odds ratio in favor of $\mathbf{H_1}$ as
$$
10^{(\textrm{TLOD} + \textrm{log prior odds})}.
$$

We show via extensive simulation that under the assumption that the mutation signature describes the biological process generating mutations in a tumor, our posterior odds ratio is a better classifier than MuTect at any threshold.
Unfortunately, this increased classification performance comes at a cost in terms of calibration, The probabilities from this model are substantially worse than those from Mutect, such that threshold selection is essentially impossible and precision/recall is substantially worse for this model.
There are several distributional assumptions in this model that effect model calibration, and we address all of them below.

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Signature & $\mu$ & AUROC &  & AUPRC & & ICI & 
\\
& & & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
\\
\colrule
100X WGS & Concentrated & Row 1 & Row 1 & -- & -- &&
\\
100X WGS & Intermediate & Row 1 & Row 1 & -- & -- &&
\\
100X WGS & Diffuse & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Concentrated & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Intermediate & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Diffuse & Row 1 & Row 1 & -- & -- &&
\\
AML 31 & Actual & Row 1 & Row 1 & -- & -- &&
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}

>>>>

Here we present an algorithm for estimating the prior probability of mutation at a given site using the observed mutation spectrum of the tumor as well as its mutation rate, and show that the addition of this prior to the MuTect variant calling model produces a superior variant classifier in both simulated and real tumor data.
We then extend the method with an application of the local false discovery rate by computing the probability that a site is non-null under an assumption of clonal expansion with either early or small selective differences between clones.
We provide a simple implementation in R that takes MuTect caller output as input, and returns the posterior probability that a site is variant for every site observed by MuTect.
\newpage
\section{MATERIALS AND METHODS}
\subsection{Somatic variant calling base probability model}

\bkmcomment{I feel like a discussion of the mutation signature should go first. Referencing the figure? The issue is that I have this m,M notation but the context C rolls in the m.}
At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
  \begin{array}{l}
    \mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
    \mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
  \end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as

Making the common assumption that NGS reads are independent, the posterior probability of a given hypothesis is 
$$
  \begin{array}{rl}
    P(m,\nu) = &p(m,\nu) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}) \\
             = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
  \end{array}
$$
where the density $\textrm{f}_{m,\nu}(x_i)$ is

$$
  \textrm{f}_{m,\nu}({b_i,q_i}) = \left\{
    \begin{array}{cr}
      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
      \frac{10^{-q_i/10}}{3} & otherwise.
    \end{array}
    \right.
$$

Assuming that the identity of the alternate allele and the estimated allele frequency are independent of each other, and $\nu$ is uniformly distributed results in the final model
$$
  P(m,\nu) = p(m) \cdot \mathcal{L}_{m,\nu}(\mathbf{x}).
$$

\subsection{Site-specific prior probability of mutation}

The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted $p(m,M | C)$(citations).
The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
$$
\mathrm{p}(m,M \mid C) = \mathrm{p}(C \mid M,m) \frac{p(m,M)}{p(C)}
$$
\bkmcomment{This is a problem. I am having trouble with the fact that C includes m. Maybe $p(C | M)$}

@@

\begin{table*}[b]
\tableparts{%
\caption{This is a table with auroc, auprc, ICI, and maybe fraction called?}
\label{table:01}%
}{%
\begin{tabular*}{\textwidth}{@{}lllllllll@{}}
\toprule
Depth & Signature & $\mu$ & AUROC &  & AUPRC & & ICI & 
\\
& & (estimated) & MuTect & Calibrated & MuTect & Calibrated & MuTect & Calibrated
\\
\colrule
100X WGS & Concentrated & Row 1 & Row 1 & -- & -- &&
\\
100X WGS & Intermediate & Row 1 & Row 1 & -- & -- &&
\\
100X WGS & Diffuse & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Concentrated & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Intermediate & Row 1 & Row 1 & -- & -- &&
\\
500X WES & Diffuse & Row 1 & Row 1 & -- & -- &&
\\
AML 31 & Actual & $3.6e-8$ & Row 1 & -- & -- &&
\\
\botrule
\end{tabular*}%
}
{$\mu$ = per-base mutation rate, AUROC/AUPRC = area under ROC/PRC curve, ICI = Integrated calibration index.}
\end{table*}

### 2019-07-17 workout_probabilities.Rmd
make a start fixing the methods/probability model section

# Somatic variant calling base statistical model

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternative non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as

$$
  \textrm{f}_{\nu,m}(b_i,q_i) = \left\{
    \begin{array}{cr}
      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
      \frac{10^{-q_i/10}}{3} & otherwise.
    \end{array}
    \right.
$$

The likelihood under the hypothesis is then $\mathcal{L}_{\nu,m}(\mathbf{x}) = \prod_{i=1}^{D} \textrm{f}_{\nu,m}(x_i)$.
MuTect reports the log likelihood ratio $\mathrm{log}(\mathcal{L}_{\nu=\hat{f},m}(\mathbf{x})/\mathcal{L}_{\nu=0,m}(\mathbf{x}))$ as either TLOD or t_lod_fstar depending on the version.
By fixing the threshold posterior odds at two, the site-specific mutation probability a constant $\mathrm{p}(M)= \mu = 3\mathrm{e}{-6}$, and $\mathrm{p}(m \mid M)$ the prior probability of mutation to specific allele $m$ constant $\mathrm{p}(m \mid M) = \mu/3 = 1\mathrm{e}{-6}$, they derive a TLOD threshold of 6.3 for classifying a site as a somatic variant.
Here we examine the effect of the assumption of a constant prior probability of mutation.

@@

>>>>

# Somatic variant calling base probability model

At every site in the genome with non-zero coverage, Next Generation Sequencing (NGS) produces a vector $\mathbf{x}  = (\{b_i\},\{q_i\}), i = 1\dots D$ of base calls and their associated quality scores, where $D$ is total read depth.
The goal is to use $\mathbf{x}$ to select between competing hypotheses;
$$
\begin{array}{l}
\mathbf{H_0}:\quad \textrm{Alt allele} = m;\quad\nu = 0\\
\mathbf{H_1}:\quad \textrm{Alt allele} = m;\quad\nu = \hat{f},
\end{array}
$$
where $\nu$ is the variant allele frequency, $\hat{f}$ is the maximum likelihood estimate of $\nu$ given data $\mathbf{x}$, i.e. the ratio of the count of variant reads and total read depth, and $m$ is any of the 3 possible alternate non-reference bases.
For a given read with base $b_i$ and q-score $q_i$, the density function under a particular hypothesis is defined as

Making the common assumption that NGS reads are independent, the posterior probability of a given hypothesis, $P(m,\nu)$ is 
$$
  \begin{array}{cr}
    P(m,\nu) = &p(m,\nu) \cdot $\mathcal{L}_{m,\nu}(\mathbf{x}). \\
             = &p(m,\nu) \cdot \prod_{i=1}^{D} \textrm{f}_{m,\nu}(x_i),
  \end{array}
$$
where the density \textrm{f}_{m,\nu}(x_i) is
$$
  \textrm{f}_{m,\nu}({b_i,q_i}) = \left\{
    \begin{array}{cr}
      \nu \frac{10^{-q_i/10}}{3} + (1-\nu)(1-10^{-q_i/10}) & b_i = \textrm{reference}\\
      \nu(1-10^{-q_i/10}) + (1-\nu) \frac{10^{-q_i/10}}{3} & b_i = m\\
      \frac{10^{-q_i/10}}{3} & otherwise.
    \end{array}
    \right.
$$

Assuming that the identity of the alternate allele and the estimated allele frequency are independent of each other, and $\nu$ is uniformly distributed results in the final model
$$
  P(m,\nu) = p(m) \times $\mathcal{L}_{m,\nu}(\mathbf{x}).
$$

@@

The probability which we have denoted $p(m)$ for compactness above is actually the joint probability that a mutation has occured $M$, and that it was to allele $m$, which we will denote here $p(m,M)$.
However, $p(m,M)$ is not constant for every site in the genome, and follows a distribution -- conditional on the genomic context surrounding the site -- described by the mutation signature of the tumor and denoted $p(m,M | C)$(citations).
The prior probability $\mathrm{p}(m,M \mid C)$ can be decomposed as
$$
\mathrm{p}(m,M \mid C) = \mathrm{p}(C \mid M,m) \frac{p(m,M)}{p(C)}
$$
\bkmcomment{This is a problem. I am having trouble with the fact that C includes m. Maybe $p(C | M)$}
